Deployable Probabilistic Programming

### 2019-04-29 paper/paper.tex
some fixes

\linespread{1.05}

@@

A probabilistic model in Infergo is an implementation of an
interface requiring a single method
\lstinline{Observe}\footnote{The method name follows
Venture~\cite{MSP14}, Anglican~\cite{TMY+16}, and other
probabilistic programming languages in which \textit{observe} is
the language construct for conditioning of the model on
observations.} accepts a vector (a Go \textit{slice}) of floats,
the parameters to infer, and returns a single float, interpreted
as unnormalized log-likelihood of the posterior distribution.
Implementation of model methods can be written in virtually
unrestricted Go and use any Go libraries.

@@

The eight schools problem~\cite{GCS+03} is an application of
the hierarchical normal model that often serves as an introductory
example of Bayesian statistical modeling. Without going into
details, in this problem the effect of coaching programs on test
outcomes is compared among eight schools.
Figure~\ref{fig:eight-schools} shows specifications of the same
model in Stan and Infergo, side-by-side.
\begin{figure*}
  \begin{minipage}{0.45\textwidth}
  \begin{lstlisting}[language=Stan,framexleftmargin=10pt,numbers=left]
data {
  int<lower=0> J;
  vector[J] y;
  vector<lower=0>[J] sigma;
}

@@

\centering
a. Stan
  \end{minipage}
  \hfill
  \begin{minipage}{0.45\textwidth}
\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
type SchoolsModel struct {
  J          int
  Y          []float64
  Sigma      []float64
}

@@

\begin{minipage}{0.49\textwidth}
\centering
\textbf{Infergo}

@@

  // Destructure parameters
  theta := make([][]float64, m.M)
  m.simplices(&x, m.K, theta)
  phi := make([][]float64, m.K)
  m.simplices(&x, m.V, phi)

@@

  // Condition on observations
  gamma := make([]float64, m.K)
  for in := 0; in != m.N; in++ {
    for ik := 0; ik != m.K; ik++ {
      gamma[ik] = math.Log(theta[m.Doc[in]-1][ik]) +
        math.Log(phi[ik][m.Word[in]-1])
    }
    ll += D.LogSumExp(gamma)
  }
  return ll
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}{0.49\textwidth}
	\centering

@@

  for (n in 1:N) {
    real gamma[K];
    for (k in 1:K) 
      gamma[k] <- log(theta[doc[n],k]) 
	  	+ log(phi[k,w[n]]);
    increment_log_prob(log_sum_exp(gamma));
  }
}
\end{lstlisting}
	\vspace{2\baselineskip}
\end{minipage}
\end{figure*}

>>>>

\linespread{1.06}

@@

A probabilistic model in Infergo is an implementation of an
interface requiring a single method
\lstinline{Observe}\footnote{The method name follows
Venture~\cite{MSP14}, Anglican~\cite{TMY+16}, and other
probabilistic programming languages in which \textit{observe} is
the language construct for conditioning of the model on
observations.} accepts a vector (a Go \textit{slice}) of floats,
the parameters to infer, and returns a single float, interpreted
as unnormalized log-likelihood of the posterior distribution.
The model object usually encapsulates the observations, either
as a data field or as a reference to a data source.
Implementation of model methods can be written in virtually
unrestricted Go and use any Go libraries.

@@

The eight schools problem~\cite{GCS+03} is an application of
the hierarchical normal model that often serves as an introductory
example of Bayesian statistical modeling. Without going into
details, in this problem the effect of coaching programs on test
outcomes is compared among eight schools.
Figure~\ref{fig:eight-schools} shows specifications of the same
model in Stan and Infergo, side-by-side.
\begin{figure*}
\begin{minipage}[t]{0.45\textwidth}
  \begin{lstlisting}[language=Stan,framexleftmargin=10pt,numbers=left]
data {
  int<lower=0> J;
  vector[J] y;
  vector<lower=0>[J] sigma;
}

@@

\centering
a. Stan
  \end{minipage}
  \hfill
\begin{minipage}[t]{0.5\textwidth}
\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
type SchoolsModel struct {
  J          int
  Y          []float64
  Sigma      []float64
}

@@

\begin{minipage}[t]{0.49\textwidth}
\centering
\textbf{Infergo}

@@

  // Regularize the parameter vector
  ll += Normal.Logps(0, 1, x...)
  // Destructure parameters
  theta := make([][]float64, m.M)
  m.FetchSimplices(&x, m.K, theta)
  phi := make([][]float64, m.K)
  m.FetchSimplices(&x, m.V, phi)

@@

  // Condition on observations
  gamma := make([]float64, m.K)
  for in := 0; in != m.N; in++ {
    for ik := 0; ik != m.K; ik++ {
      gamma[ik] = math.Log(theta[m.Doc[in]-1][ik]) +
        math.Log(phi[ik][m.Word[in]-1])
    }
    ll += D.LogSumExp(gamma)
  }
  return ll
}

func (m *Model) FetchSimplices(
	px *[]float64,
	k int,
	simplices [][]float64,
) {
	for i := range simplices {
		simplices[i] = make([]float64, k)
		D.SoftMax(model.Shift(px, k), simplices[i])
	}
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}[t]{0.49\textwidth}
	\centering

@@

  for (n in 1:N) {
    real gamma[K];
    for (k in 1:K) 
      gamma[k] <- log(theta[doc[n],k]) 
	  	+ log(phi[k,w[n]]);
    increment_log_prob(log_sum_exp(gamma));
  }
}
\end{lstlisting}
\end{minipage}
\end{figure*}
